---
title: "simplified_gbcd_with_diagonal_analysis"
author: "Annie Xie"
date: "2025-10-08"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

The aim of this vignette is to explore if the diagonal component affects the resulting decomposition found by the "simplified gbcd" procedure.

We begin by loading the necessary packages. We also set the seed so that results can be fully reproduced.    
```{r load_packages, message = FALSE, warning = FALSE}
library(flashier)
library(ashr)
library(ebnm)
library(Matrix)
library(pheatmap)
library(ggplot2)
library(fastTopics)
library(cowplot)
library(dplyr)
set.seed(3)
```

# 1000 Genomes Project Data

We load in the scaled Gram matrix.
```{r load_data}
S <- readRDS('data/tgp_cov_matrix.rds')
```

We also load in the meta data which contains population information for each sample.
```{r load_meta}
tgp_meta <- readRDS('data/tgp_meta.rds')
```

# Using Flashier to fit GBCD
Now, we use the R package `flashier` to apply a simplified version of GBCD on the 1000 Genomes Project data. This is broken down into several steps.

Step 1: Fit a point-Laplace EBMF fit to the scaled Gram matrix

```{r}
cov_fit <- function(covmat, ebnm_fn = ebnm::ebnm_point_laplace, Kmax = 1000, verbose.lvl = 0, backfit = TRUE) {
  fl <- flash_init(covmat, var_type = 0) %>%
    flash_set_verbose(verbose.lvl) %>%
    flash_greedy(ebnm_fn = ebnm_fn, Kmax = Kmax)
  if (backfit == TRUE){
    fl <- flash_backfit(fl)
  }
  s2 <- max(0, mean(diag(covmat) - diag(fitted(fl))))
  s2_diff <- Inf
  while(s2 > 0 && abs(s2_diff - 1) > 1e-4) {
    covmat_minuss2 <- covmat - diag(rep(s2, ncol(covmat)))
    fl <- fl %>% 
      flash_update_data(covmat_minuss2) %>%
      flash_set_verbose(verbose.lvl) %>%
      flash_backfit() %>%
      flash_nullcheck()
    old_s2 <- s2
    s2 <- max(0, mean(diag(covmat) - diag(fitted(fl))))
    s2_diff <- s2 / old_s2
  }
  
  return(list(fl=fl, s2 = s2))
}
```

```{r pt_laplace_fit}
tgp_cov_fit <- cov_fit(covmat = S, 
                       ebnm_fn = ebnm::ebnm_point_laplace, 
                       Kmax = 11)
```

Step 2: Split the factors into positive and negative components and refit the weights

```{r split_function}
split_pt_laplace_estimates <- function(cov_fit){
  # S is the scaled Gram matrix
  # cov_fit is a flash object
  
  LL <- cov_fit$L_pm
  FF <- cov_fit$F_pm
  
# split into positive and negative components
  LL <- cbind(pmax(LL, 0), pmax(-LL, 0))
  FF <- cbind(pmax(FF, 0), pmax(-FF, 0))
  
  # remove columns of zeros
  LL.idx.nonzero <- apply(LL, 2, function(x){return(sum(x^2))}) > 10^(-10)
  FF.idx.nonzero <- apply(FF, 2, function(x){return(sum(x^2))}) > 10^(-10)
  LL <- LL[, LL.idx.nonzero]
  FF <- FF[, FF.idx.nonzero]
  
  # refit weights by least squares
  # n <- nrow(S)
  # lft_vec <- matrix(rep(0, ncol(LL)*n*n), ncol = ncol(LL))
  # for (i in 1:ncol(LL)){
  #   lft_vec[,i] <- c(LL[,i]%*%t(FF[,i]))
  # }
  # 
  # nnlm_fit <- NNLM::nnlm(lft_vec, as.matrix(c(S), ncol = 1), alpha = c(0,0,0))
  # 
  # indices_keep <- (nnlm_fit$coefficients > 0)
  # LL_scaled <- LL[,indices_keep] %*% diag(sqrt(nnlm_fit$coefficients[indices_keep]))
  # FF_scaled <- FF[,indices_keep] %*% diag(sqrt(nnlm_fit$coefficients[indices_keep]))
  return(list(LL, FF))
}
```

```{r pt_laplace_split_init}
pt_laplace_split_initialization <- split_pt_laplace_estimates(tgp_cov_fit$fl)
```

Step 3: Fit a generalized-binary EBMF fit to the scaled Gram matrix

```{r}
nonnegative_cov_fit <- flash_init(S) |>
  flash_factors_init(pt_laplace_split_initialization, ebnm_fn = ebnm::ebnm_generalized_binary) |>
  flash_backfit(maxiter = 25) |>
  flash_nullcheck()
```

```{r}
cov_fit_backfit <- function(covmat, init_LF_list, ebnm_fn = ebnm::ebnm_point_laplace, verbose.lvl = 2, backfit_maxiter = 200, maxiter = 100) {
  fl <- flash_init(covmat, var_type = 0) |>
    flash_set_verbose(verbose.lvl) |>
    flash_factors_init(init_LF_list, ebnm_fn = ebnm_fn)
  s2 <- max(0, mean(diag(covmat) - diag(fitted(fl))))
  s2_diff <- Inf
  iter <- 1
  while(s2 > 0 && abs(s2_diff - 1) > 1e-4 && iter <= maxiter) {
    covmat_minuss2 <- covmat - diag(rep(s2, ncol(covmat)))
    fl <- fl |>
      flash_update_data(covmat_minuss2) |>
      flash_set_verbose(verbose.lvl) |>
      flash_backfit(maxiter = backfit_maxiter) |>
      flash_nullcheck()
    old_s2 <- s2
    s2 <- max(0, mean(diag(covmat) - diag(fitted(fl))))
    s2_diff <- s2 / old_s2
    iter <- iter + 1
  }
  
  return(list(fl=fl, s2 = s2))
}
```

```{r gb_fit}
tgp_cov_gb_fit <- cov_fit_backfit(covmat = S, list(nonnegative_cov_fit$L_pm, nonnegative_cov_fit$F_pm),
                               ebnm_fn = ebnm::ebnm_generalized_binary, maxiter = 1000)
```

```{r scaled_gb_fit}
tgp_cov_fit_scaled <- ldf(tgp_cov_gb_fit$fl, 'i')
```

Step 4: Filter factors that are not similar across both estimates

```{r scaled_loadings}
tgp_cov_fit_scaled_L <- tgp_cov_fit_scaled$L %*% diag(sqrt(tgp_cov_fit_scaled$D))
tgp_cov_fit_scaled_F <- tgp_cov_fit_scaled$F %*% diag(sqrt(tgp_cov_fit_scaled$D))
```

```{r filter_factors}
# compute correlations
factor_correlations <- diag(cor(tgp_cov_fit_scaled_L, tgp_cov_fit_scaled_F))
hist(factor_correlations, main = 'Histogram of Correlations', xlab = 'Correlations')
```

```{r}
factor_keep <- factor_correlations > 0.8
```

```{r final_loadings_estimate}
tgp_cov_fit_filtered_scaled_L <- tgp_cov_fit_scaled_L[, factor_keep]
```

# Visualizing the Results

A useful plot type for visualizing non-negative loadings matrices is the "structure plot" -- a stacked bar plot in which each component is represented as a bar of a different color, and the bar heights are given by the loadings on each component. 

This is the structure plot for the loadings estimate from gbcd:
```{r structure_plot_L}

factor_colors <- Polychrome::kelly.colors(ncol(tgp_cov_fit_filtered_scaled_L)+1)[-1]

L <- tgp_cov_fit_filtered_scaled_L
k <- ncol(L)
colnames(L) <- paste0("k",1:k)
p1 <- structure_plot(L,grouping = tgp_meta$pop,
                     colors = factor_colors,
                     gap = 16,verbose = FALSE) +
  labs(y = "membership",title = "GBCD")
p1
```

Factor 1 (“k1” in the legend) appears to be a "baseline" factor. To focus on differences between samples, we can remove first factor from our structure plot:

```{r structure_plot_wout_k1}
p2 <- structure_plot(L[,-1],grouping = tgp_meta$pop,
                     colors = factor_colors[-1],
                     gap = 16,verbose = FALSE) +
  labs(y = "membership",title = "GBCD")
p2
```

From this plot, we can interpret the components of our decomposition. Some of the factors primarily capture super-population effects; factor 2 primarily captures the Africa super-population, factors 4 and 6 primarily capture the East Asia super-population, factor 7 primarily captures the South Asia super-population, and factor 8 primarily captures the Americas super-population. Factor 3 captures shared structure between the Europe super-population and the Americas super-population. Factor 5 captures shared structure between two African populations, Mende in Sierra Leone and Gambian in Western Division - Mandinka. Factor 10 captures shared structure between three European populations -- British from England and Scotland (GBR), Northern Europeans from Utah (CEU), and Finnish in Finland (FIN). 

Some of the factors capture population-specific effects. Factor 9 corresponds to Japanese in Tokyo, Japan (JPT), differentiating it from the other East Asian populations. Furthermore, factor 12 corresponds to a unique component for the Luhya people of Webuye, Kenya.

The results also highlight interesting compositions for particular populations. For example, Peruvian from Lima, Peru (PEL) contains one component, factor 8, which primarily corresponds to the Americas super-population. However, it also contains other components, factors 4 and 6, which are shared with South Asian and/or East Asian populations. Furthermore, it contains a component, factor 3, that is shared with Europe and other populations in the Americas super-population.
